---
title: "create_fullaseg_atlas.R"
output: html_document
date: '2023-11-16'
---

Creating new atlas based on aseg to plot global tissue phenotypes. Code based on the example [here](https://ggseg.github.io/ggsegExtra/articles/createaseg.html).

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
if(!require('pacman')) {
  install.packages('pacman')
}
pacman::p_load(tidyverse, dplyr)
```

## SETUP
Code requires that you have [FSL](https://fsl.fmrib.ox.ac.uk/fsl/fslwiki/FslInstallation), [FreeSurfer](https://surfer.nmr.mgh.harvard.edu/fswiki/DownloadAndInstall) (with `license.txt`), [ImageMagick](https://imagemagick.org/script/download.php), and [orca](https://github.com/plotly/orca) installed.

Install/load R packages as needed:
*If R asks whether you want to install packages that require compillation, say no!*

### freesurfer
This package has been archived from Cran, but can be installed as follows:

```{r}
require(devtools)
install_version("freesurfer", version = "1.6.8", repos = "http://cran.us.r-project.org")

Sys.setenv(FREESURFER_HOME = "/Applications/freesurfer/7.4.1") #point to freesurfer install
```

### ggsegExtra
can also be tricky to install! should load ggseg() and ggseg3D automatically.
```{r}
options(repos = c(
  ggseg = 'https://ggseg.r-universe.dev',
  CRAN = 'https://cloud.r-project.org'))
install.packages('ggsegExtra')
```
Load
```{r}
library(freesurfer)
library(ggsegExtra)
```

## MAKE ATLAS

Create atlas from freesurfer average output
**update paths as needed**

```{r}
aseg_all <- ggsegExtra::make_volumetric_ggseg(subject = "fsaverage5",
                               steps = 1:8,
                               output_dir = "~/Desktop/BGD_Repos/ggseg_atlas", 
                               vertex_size_limits = c(10, NA),
                               label_file = file.path("/Applications/freesurfer/7.4.1/subjects/fsaverage5/mri/aseg.mgz"))

aseg_all$data <- filter(aseg_all$data, 
                     !grepl("Unknown", label, ignore.case = TRUE)) #drop "unknown" region
plot(aseg_all, alpha = .8)
```


```{r}
#add col for tissue class
aseg_all$data <- aseg_all$data %>%
  mutate(tissue_class = as.factor(case_when(grepl("ventricle", label, ignore.case = TRUE) ~ "CSF",
                                            grepl("choroid", label, ignore.case = TRUE) ~ "CSF",
                                            grepl("cerebellum", label, ignore.case = TRUE) ~ "Cerebellum",
                                            grepl("white", label, ignore.case = TRUE) ~ "WM",
                                            grepl("CSF", label, ignore.case = TRUE) ~ "CSF",
                                            grepl("cerebral", label, ignore.case = TRUE) ~ "cGM",
                                            grepl("stem", label, ignore.case = TRUE) ~ "Brainstem",
                                            TRUE ~ "sGM")))
ggplot() +
  ggseg::geom_brain(atlas=aseg_all, aes(fill = tissue_class))
```

Save out full atlas

```{r}
saveRDS(aseg_all, file="aseg_all_atlas.RDS")
```

Create second version excluding cerebellum and brainstem

```{r}
#make reduced version that doesn't include cerebellum and brainstem
aseg_cerebrum <- aseg_all
aseg_cerebrum$data <- aseg_cerebrum$data %>%
  dplyr::filter(tissue_class != "Cerebellum" & tissue_class != "Brainstem")

saveRDS(aseg_cerebrum, file="aseg_cerebrum_atlas.RDS")

ggplot() +
  ggseg::geom_brain(atlas=aseg_cerebrum, aes(fill = tissue_class))
```

